---
title: "Uncertainty of models and scenarios"
output: html_document
editor_options: 
  chunk_output_type: console
---

So far we have looked at relative severity of single units of observation (raster cells or glacier outlines) but we are interested in describing degradation of abiotic conditions in the whole assessment unit.

```{r}
#| eval: true
#| echo: false
#| message: false
library(dplyr)
library(ggplot2)
library(readr)
library(ggforce)
library(DT)
library(htmltools)
library(purrr)
library(ggpubr)
library(tidyr)
library(stringr)
here::i_am("docs-src/RS/spatial-variability.qmd")
# sessionInfo()
# Suppress summarise info
options(dplyr.summarise.inform = FALSE)
source(here::here("inc","R","RS-functions.R"))
source(here::here("inc","R","RS-shared.R"))
```

```{r}
#| eval: true
#| message: false
target.dir <- "sandbox"
results_file <- here::here(target.dir, "relative-severity-degradation-suitability-all-tropical-glaciers.csv")
RS_results <- read_csv(results_file, show_col_types = FALSE) %>% 
  mutate(unit_name=str_replace_all(unit,"-"," ")) %>%
  mutate(unit_name = factor(unit_name, unit_order) )
results_file <- here::here(target.dir, "massbalance-model-data-all-groups.rds")
massbalance_results <- readRDS(results_file)

rds.file <- here::here(target.dir,"totalmass-suitability-cED-data.rds")
cED_model_data <- readRDS(rds.file)
```

## Considerations before aggregation

Our two indicator variable differ fundamentally in units and interpretation and we have to consider this when we aggregate the values of RS calculated from the sampling or observation units to larger assessment units. We also have to consider the choice of formula to calculate RS and the interpretation of the collapse threshold in each case.

### Sampling units

For our two indicator variables we have different sample sizes per assessment unit. In the case of the dynamic ice mass balance model the units of observation are glacier outlines from the Randolph Glacier Inventory, while in the bioclimatic suitability model the units of observation are cell in a spatial grid that overlap with glacier outlines from multiple sources. 

```{r}
rds.file <- here::here(target.dir,"massbalance-totalmass-all-groups.rds")
totalmass_year_data <- readRDS(rds.file) 

massbalance_results %>% 
  group_by(unit_name) %>% 
  summarise(`Sampling size (ice)` = n_distinct(RGIId)) %>%
  full_join({
    RS_results %>% 
    group_by(unit_name) %>% 
    summarise(`Sampling size (bcs)` = n_distinct(id,cellnr))
    },
    by=c("unit_name")
  )  %>% 
  datatable(
    options = list(
        dom = 't', 
        pagination = FALSE, 
        pageLength = -1
        )
    ) 

```

Differences in sample size are noticeable in one assessment units that covers areas of Peru and Chile.  The national glacier inventory of Chile include more detail about different ice accumulations types, including rock glaciers, that are not represented in the global inventories.



### Interpretation of values and collapse thresholds

The indicator variable of ice mass is a positive variable and represents a physical measurement of mass. For this variable the  collapse threshold of $0 kg$ indicates complete loss of the icy substrate. 

For the indicator variable of bioclimatic suitability we are using a probability estimated by the model that requires additional interpretation. We use a confusion matrix to calculate predictive performance of the model and decide the best cut-off value between suitable and unsuitable habitat, and use that cut-off value as an estimate of collapse threshold indicating a shift in suitability conditions.

## RS for the loss of ice mass




```{r}
#| label: fig-RS-timeseries
#| fig-cap: "Time series of relative severity (RS) of decline in total ice mass for different assessment units (columns) and three future scenarios (rows) based on a collapse threshold of $0 kg$ of ice. The points represents values calculated for 12 different global circulation models."
#| 
# alternative with purrr:
#RSvals <- totalmass_year_data %>% 
#  filter(unit_name %in% "Cordillera de Merida") %>%
#  group_by(model_nr,scn) %>% 
#  group_split() %>%
#  map(~RSts(.x$year,.x$total_mass)) %>% 
#  bind_rows()

RSvals <- totalmass_year_data %>% 
  filter(unit_name %in% c("Ruwenzori",
                          "Kilimanjaro", 
                          "Mount Kenia")) %>%
#  filter(unit_name %in% c("Ecuador")) %>%
  group_by(unit_name,model_nr,ssp) %>% 
  group_modify(~RSts(
    .x$year,
    .x$total_mass,
    vmin=.x$min_mass,
    vmax=.x$max_mass,
    formula = "conditional"
    )
    ) 

ggplot(RSvals) +
  ## geom error bar is small compared to intermodel variability, keep out of the plot
  #  geom_errorbar(aes(x=year,ymin=RS_min,ymax=RS_max),alpha=.25) +
  geom_point(aes(x=year,y=RS,colour=ssp),alpha=.50,cex=.5) +
  facet_grid(ssp~unit_name) +
  theme(legend.position = "none")

```

```{r}
jj <- "Cordilleras Norte de Peru"
ss <- "SSP2-4.5"
mm <- " 9"
mbdata <- massbalance_results %>% 
    filter(
        unit_name %in% jj,
        ssp %in% ss,
        model_nr %in% mm,
        year %in% c(2000,2050)
        ) 

    wgs <- mbdata %>% 
        filter(year == 2000) %>% 
        group_by( RGIId) %>% 
        summarise(initial_mass=sum(mass), .groups="keep") 

    RSi_data <- mbdata %>% 
    group_by(RGIId) %>% 
    group_modify(~RSts(.x$year,.x$mass,
                        formula = "conditional")) %>%
    ungroup %>% 
    left_join(wgs, by = c("RGIId"))

ggplot(RSi_data) + 
  geom_point(aes(x=IV,y=RS,colour=RS,size=FV)) + 
  scale_x_continuous(trans="log") + 
  scale_colour_continuous(trans="reverse") + 
  theme(legend.position = "none")

#library(forcats)
# RSi_data <- RSi_data %>% mutate(oo=fct_reorder(RGIId,IV))
# ggplot(RSi_data) + 
#  geom_col(aes(x=oo,y=OD,colour=RS)) 

#ggplot(RSi_data) + 
#     geom_col(aes(x=oo,y=IV),colour="pink") + 
# geom_col(aes(x=oo,y=FV),colour="slateblue4")

```

```{r}
dat1 <- RS_results %>% 
  filter(
    threshold == "acc",
    pathway == "ssp370",
    modelname == "mri-esm2-0") %>% 
    mutate(timeframe=str_replace(timeframe,"-","\n"))
dat2 <- dat1 %>% 
  group_by(timeframe,unit_name) %>% 
  summarise(mean_RS=mean(RS_cor))

dat3 <- dat1 %>% 
  group_by(unit_name,timeframe) %>% 
  group_modify(~summary_cED_w(.x$RS_cor)) %>%
  inner_join(dat2, by=c("unit_name","timeframe")) %>%
  mutate(state = case_when(
            mean_RS == 1 ~ "collapsed",
            cED_80>=0.80 ~ "very wide",
            cED_80>=0.50 ~ "very inter",
            cED_50>=0.80 ~ "high wide",
            cED_80>=0.30 ~ "very local",
            cED_50>=0.50 ~ "high inter",
            cED_30>=0.80 ~ "mod wide",
            TRUE ~ "low")
  )


```

```{r}
sts <- dat3 %>% select(unit_name,timeframe,state)
dat1 <- dat1 %>% 
  left_join(sts, by=c("unit_name","timeframe"))

  ggplot(dat1 ) +
  geom_boxplot(aes(y = RS_cor, x = timeframe),colour="grey77") +
# mucho carnaval:  
## geom_boxplot(aes(y = RS_cor, x = timeframe, colour=state))+#,colour="grey77") +
##   scale_colour_manual(values=state_cat_okabe) 
  geom_point(data=dat2, aes(y = mean_RS, x = timeframe),pch=1,cex=3,colour="grey22") +
  facet_wrap(~unit_name) +
  theme(legend.position = "none") +
  ylab(expression(RS[i] * phantom(m) * textstyle(or) * phantom(m) * bar(RS))) +
  xlab("Future period") 
```


```{r}
dat3 %>% pivot_wider(id_cols = unit_name,values_from=state,names_from=timeframe)
```
