---
title: "Temporal change in Relative Severity"
output: html_document
editor_options: 
  chunk_output_type: console
---


We calculate temporal change in Relative Severity by applying the $\mathrm{RS}_{i}$ formula to one fixed initial value and a sequence of final values, the final values can represent the state of the indicator in consecutive years (time series), or in different periods (temporal sequence). 

```{r}
#| label: load libraries and functions
#| eval: true
#| echo: false
#| message: false
library(dplyr)
library(ggplot2)
library(units)
library(stringr)
library(ggrepel)
library(purrr)
library(readr)
library(tidyr)
library(forcats)

here::i_am("docs-src/RS/time-series.qmd")
source(here::here("inc","R","RS-functions.R"))
```

```{r}
#| label: read input data 
#| eval: true
#| message: false
target.dir <- "sandbox"
rds.file <- here::here(target.dir,"massbalance-totalmass-all-groups.rds")
totalmass_year_data <- readRDS(rds.file)
results_file <- here::here(target.dir, "relative-severity-degradation-suitability-all-tropical-glaciers.csv")
RS_results <- read_csv(results_file, show_col_types = FALSE) %>%
  mutate(unit_name=str_replace_all(unit," ","-"))
```

```{r}
#| label: ggplot theme
old <- theme_set(theme_linedraw())
theme_update(panel.grid.minor = element_line(colour = "pink"),
panel.grid.major = element_line(colour = "rosybrown3"))
okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

```

## Loss of ice mass

The data from the Dynamic Ice Mass Balance model is provided in one year intervals, and we have summarise the data to provide the total value of ice mass of each assessment unit for each year.

We calculate RS using the predicted ice mass of the year 2000 as initial value, and data from subsequent years (2001 to 2100) as final values, and a collapse threshold of $0 kg$. [We denote this estimate of relative severity as $RS_{ice}^{CT=0}$.]{.aside} Given that the collapse threshold is also the minimum value of ice possible, and that increases in ice mass are low in magnitude and limited to the start of the time series, we can use the original RS formula.

An example is given in @fig-rs-ice-example-kili.

### Average RS time series

Using the equation for average RS, we can create time series of $\overline{\mathrm{RS}}$ for each assessment unit.

```{r}
#| label: fig-rs-ice-all-units
#| fig-cap: "Time series of average RS values for the ice mass indicator variable for all assessment units. We use here data from the 100 years time series calculated for one scenario (SSP3-7.0), and combined median and median absolute deviation (based on 50 replicates per model) for all 12 models. The line represent the best estimate and the shaded area represent the interval between aggregated  min and max values for each year."
#| eval: true
RStotal <- totalmass_year_data %>% 
  filter(
    scn %in% c("SSP3-7.0")
  ) %>%
  group_by(unit_name, model_nr) %>%
  group_modify(~RSts(
    .x$year,
    .x$total_mass,
    vmin=.x$min_mass,
    vmax=.x$max_mass,
    formula = "original"
    )
    ) 
RStotal <- RStotal %>%
 pivot_longer(
  cols = starts_with("RS"),
    names_to = "type",
    values_to = "RS"
 )  %>% 
 group_by(unit_name,year) %>%
 summarise(
  RSmin = min(RS),
  RSmax = max(RS),
  RSmean = mean(RS),
  .groups="keep"
  ) %>%
 mutate(
  unit_name = factor(unit_name, unit_order),
  collapsed = RSmin>0.99
  )

ggplot(RStotal) +
  geom_ribbon(aes(x = year, ymin = RSmin, ymax = RSmax), alpha=.55, fill = okabe[2]) +
  geom_line(aes(x = year, y = RSmean, colour = collapsed), alpha=.85, linewidth=1.2) +
  facet_wrap(~ unit_name, labeller=labeller(unit_name=label_wrap_gen())) +
  scale_colour_discrete(type=okabe[c(5,6)]) +
  theme(legend.position="none") +
  ylab(expression(bar(RS))) +
  xlab("Year")
  
```



## Degradation of bioclimatic suitability



  Given that both initial and final values of suitability can be below these thresholds, and that future suitability predictions can be well above current values, we use the conditional RS formula for these estimates of RS.


```{r}
#| label: fig-RS-periods
#| fig-cap: "Mean relative severity (RS) of decline in bioclimatic suitability for the assessment unit of the T.G.E. of Ecuador for five global circulation models (points), and three different timeframes (columns), future scenarios (rows) and collapse thresholds (x axis) based on threshold of the confusion matrix ( $acc$: maximum accuracy, $ess$: equal sensitivity and specificity, and $ppv$: Positive predictive value)."
#| warning: false
slc_unit <- "Ecuador"
dat1 <- RS_results %>% 
  filter(unit %in% slc_unit) %>% 
  mutate(year=case_when(
    timeframe %in% "2011-2040"~2025,
    timeframe %in% "2041-2070"~2055,
    timeframe %in% "2071-2100"~2085
    ),
    scn=str_replace(pathway,"[ssp]+([0-9])([0-9])([0-9])","SSP\\1-\\2.\\3"),) %>%
  group_by(scn,threshold,timeframe,year,modelname) %>%
  summarise(n=n(),RSmean=mean(RS_cor),RSmed=median(RS_cor),
  .groups="keep")

ggplot(dat1) +
  geom_point(data=dat1,
               aes(y = RSmean, 
                   colour=threshold, 
                   x=timeframe)) + 
  facet_grid(scn~threshold) +
  ylab("Mean RS") +
  xlab("Collapse thresholds") +
  theme(legend.position="none")
```

----

<details><summary>R session info for this chapter </summary>

```{r}
sessionInfo()
```
</details>