---
title: "Ecosystem description"
output: html_document
editor_options: 
  chunk_output_type: console
---

Tropical Glacier Ecosystems are cryogenic ecosystems in tropical environments where the main substrate is formed by ice accumulated over several years and provides different habitats to microbiological communities and other meso- and macro-biota.

Tropical Glacier Ecosystems are originally formed by processes of snow and ice accumulation and ablation over geological time frames. However the characteristic microbiota can persist in the ice substrate even if processes of ice accumulation have ceased or the overall mass balance of ice is negative.

```{r}
#| warning: false
#| message: false
library(sf)
require(leaflet)
library(dplyr)
library(ggplot2)

here::i_am("docs-src/TGE/units.qmd")
target.dir <- "sandbox"
trop_glaz <- read_sf(here::here(target.dir,"trop-glacier-groups-labelled.gpkg"))
load(here::here(target.dir,"current-bioclim-data-all-groups.rda"))
```

```{r}
#| eval: false

## This code checks if the ids of the groups match

input_bioclim_data <-
    st_as_sf(
        input_raster_data, 
        coords = c("lon", "lat"), 
        crs = 4326
    )
tst <- st_intersection(trop_glaz, slice_sample(input_bioclim_data,n=100)) 
tst %>% transmute(test=id==id.1) %>% st_drop_geometry %>% pull %>% table
```



## Classification

We use the IUCN Global Ecosystem Typology [@Keith_2020_Ecosystem_Typology] as a reference ecosystem classification system and identify ***Tropical Glacier Ecosystems*** as a subgroup within the ecosystem functional group _T6.1 Ice sheets, glaciers and perennial snowfields_:

IUCN Global Ecosystem Typology

- Level 1: Terrestrial Realm
  - Level 2: T6 Polar/Alpine (cryogenic) functional biome
    - Level 3: T6.1 Ice sheets, glaciers and perennial snowfields
      - Subgroup: Tropical Glacier Ecosystems
      - Level 4: 12 units as follows:

```{r}
#| label: name of units

exclude <- c("Temperate Glacier Ecosystems", "Famatina", "Norte de Argentina", "Zona Volcanica Central")

trop_glaz_names <- trop_glaz %>% 
  st_drop_geometry()  %>% 
  filter(!group_name %in% exclude) %>%
  arrange(group_name) %>%
  pull(group_name) %>%
  unique 
sprintf("T.G.E. %s", trop_glaz_names)
```

We exclude transitional units

## Spatial distribution

We use glacier outlines

The Tropical Glacier Ecosystem are found in the highest tropical mountains above the snowline. 

We use available global and regional data on glacier outlines and point records to estimate the global distribution of Tropical Glacier Ecosystems. In the Randolph Glacier Inventory version 6.0 [@RGI_2017_glacier_inventory] all tropical glaciers are includes in region ... Lower Latitudes. We compared these records with other global sources [] and include additional point records from national glacier inventories of Argentina and Chile to increase the coverage of these regions [@fig-map]. 

```{r}
#| label: fig-map
#| fig-cap: Outline of spatial assessment units for Tropical Glacier Ecosystems. Each assessment unit consist of one or more spatial clusters (polygons), and each cluster includes a buffer of 25km around the glacier outlines. 
#| warning: false
#| eval: true
t_lbls <- sprintf("<strong>%s</strong> (%s)",trop_glaz$group_name,trop_glaz$id) %>% lapply(htmltools::HTML)

leaflet() %>%
  addProviderTiles(providers$OpenStreetMap.HOT,group='OpenStreetMap') %>%
  addProviderTiles(providers$OpenTopoMap,group='OpenTopoMap') %>%
  addProviderTiles(providers$Esri.WorldImagery,group='Esri WorldImagery') %>%
  addPolygons(data = trop_glaz , fillColor = "plum", color='magenta',
              weight = 0.8, fillOpacity = 0.15, label=t_lbls,
              highlightOptions = highlightOptions(weight = 2, color = 'orange'),
              group="T.G.E. units")  %>%
  addLayersControl(
    baseGroups = c('OpenStreetMap','OpenTopoMap','Esri WorldImagery'),
    overlayGroups = c("T.G.E. units"),
    options = layersControlOptions(collapsed = FALSE),
    position = "topright"
  )  %>% hideGroup("Spatial buffer")
```

## Characteristic biota

Tropical Glacier Ecosystems have five main ecological zones [@Hotaling_2017_glacier_ecosystem]: the supraglacial surfaces, englacial interior, subglacial bedrock–ice interface, proglacial streams and lakes and glacier forefields. 

## Definition of collapsed states

```{r out.height="400px"}
#| label: fig-CEM
#| fig-cap: Simplified Conceptual Ecosystem Model for Tropical Glacier Ecosystems
DiagrammeR::grViz("CEM-all-TGE.gv")
```