---
title: "Prediction of collapse"
output: html_document
editor_options: 
  chunk_output_type: console
---

Over the time frame of 100 years most of the assessment units are predicted to reach collapse by complete loss of ice mass. Since we are using a direct indicator of an ecosystem property (icy substrate) and we are predicting total ice mass for each unit, complete loss of ice is equivalent to a value of $RS_{ice}^{CT=0} = 1$. In the case of an indirect indicator such as bioclimatic suitability, we have more uncertainty in the real value of collapse, and thus use alternative collapse threshold to capture plausible ranges. 

We expect that both direct and indirect indicators will have similar performance in describing the magnitude of degradation and predicting collapse. We will use two generalised linear mixed models (GLMM) to test if there are significant differences in inferences based on these estimates of relative severity.


```{r}
#| label: load libraries and functions
#| eval: true
#| echo: false
#| warning: false
#| message: false
library(dplyr)
library(ggplot2)
library(units)
library(stringr)
library(ggrepel)
library(purrr)
library(readr)
library(tidyr)
library(glmmTMB)
library(lme4)
library(DHARMa)
here::i_am("docs-src/GLMM/binomial-collapse.qmd")
source(here::here("inc","R","RS-functions.R"))
source(here::here("inc","R","RS-shared.R"))
```

```{r}
#| label: read input data 
#| eval: true
#| message: false
target_dir <- "sandbox"
rds_file <- here::here(target_dir,"massbalance-totalmass-all-groups.rds")
totalmass_year_data <- readRDS(rds_file)
results_file <- here::here(target_dir, "relative-severity-degradation-suitability-all-tropical-glaciers.csv")
RS_results <- read_csv(results_file, show_col_types = FALSE) %>%
  mutate(unit_name=str_replace_all(unit," ","-"))
rds_file <- here::here(target_dir,"totalmass-suitability-glmm-data.rds")
model_data <- readRDS(rds_file)

```


To compare both indicators we used the $RS_{bcs}^{CT=acc}$ and $RS_{bcs}^{CT=ess}$ for the future periods paired with the $RS_{ice}^{CT=0}$ values for the years 2040, 2070 and 2100. We coded these three periods/years as variable $\mathrm{time}$ with values $0, 1, 2$ respectively. We use the respective total or mean $RS$ values to calculate the response variable and includes a categorical variable $\mathrm{method}$ with three levels indicating either the direct indicator ($ice$) or indirect indicator with two alternative thresholds ($acc$ or $ess$).

## Proportion of models predicting collapse


```{r}
valid_methods <- c("ice","ess","acc")
model_data <- model_data %>%
  filter(method %in% valid_methods) %>%
  mutate(
    method=droplevels(method),
    collapsed=case_when(
      method %in% "ice" & RS == 1 ~ 1L,
      !(method %in% "ice") & RS >0.99 ~ 1L,
      TRUE ~ 0L
    )
    )
model_data %>%
  group_by(unit,method) %>%
  summarise(collapse=sum(collapsed,na.rm=T), .groups = "keep") %>%
  pivot_wider(values_from=collapse,names_from=method)

```


## Binomial GLMM of predicted collapse

We used a binomial GLMM with logit link function, using a response variable with values $y=1$ when $RS=1$ and $y=0$ otherwise. We included fixed effects of scenarios ($\mathrm{scenario}$ with three levels) and $\mathrm{time}$, and rabdom effects of assessment unit ($\mathrm{unit}$, 15 levels). [Each observation corresponds to the prediction of one global circulation model, but since models are not identified in the ice mass balance model, we threat the different models as anonymous replicates, and this implies that the effect of the model is nested within $\mathrm{method}$.]{.aside}

The variable $\mathrm{method}$ could be interpreted as a fixed effect and/or as a random effect *grouping* variable. Although it might be interesting to explore its interaction with unit in increasing the variability of the response [@Barr_2013_GLMM_effects], our primary question is whether there are significant systematic differences between the methods. So we decide to use this variable as a fixed effect and keep the model simple for interpretability:

Full model specification in `R` using the `glmmTMB` package is as follows:

```{r}
#| echo: true
mod_collapse_fixed <- 
  glmmTMB(collapsed ~ time + method + scenario + (1|unit), 
       data = model_data, 
       family = binomial,
       REML = FALSE)
```

::: {.column-margin}
```{r}
#| echo: true
#| eval: false

## alternative with lme4
library(lme4)
mod_collapse_full <- glmer(
    collapsed ~ time + method + scenario + (1|unit), 
    family = binomial,
    data = model_data) 
```
:::



Alternative with both fixed and random, will have similar fixed effects, but could be used to measure the amount of variability attributed to methods vs. units.
```{r}
#| echo: true
#| eval: true

mod_collapse_both <- 
  glmmTMB(collapsed ~ time + scenario + method + (1|unit/method), 
     data = model_data, 
     family=binomial,
     REML=FALSE)
```

The AIC criterion clearly favours the model with methods as a fixed and a random grouping variable:

```{r}
#| echo: true
bbmle::AICtab(mod_collapse_fixed,
              mod_collapse_both)
```

### Model diagnostics

Model diagnostics and residual plots look good for this model specification. [According to the [vignette of package DHARMa](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html#interpreting-residuals-and-recognizing-misspecification-problems) small significant effects in the residuals might be spurious due to large sample size.]{.aside}

```{r}
#| echo: true
mod_collapse_simres<-simulateResiduals(mod_collapse_both)
plot(mod_collapse_simres)
```
 
  
### Explained variance

Approximation of a $R^2$ statistic for this model suggests more than 90% of the variance explained by the full model (random and fixed effects) and at least 40% explained by the fixed effects alone.

```{r}
#| echo: true
#| warning: false
MuMIn::r.squaredGLMM(mod_collapse_both)
```


### Model summary

The summary of the model indicates significant positive effects of time and future scenarios in the proportion of model predicting collapse, as expected. For the method variable, the indirect indicator have negative effects when compared with the direct indicator, but this is only significant for the maximum accuracy threshold. We can interpret this to be the lower, more conservative or optimistic bound of the collapse threshold for this indicator. 

Random effect of unit is larger than the random effect of methods within units.

```{r}
#| echo: true
summary(mod_collapse_both)
```

```{r}
#| echo: true
summary(aov(mod_collapse_both))
```


Confidence interval of the coefficients:
```{r}
#| echo: true
confint(mod_collapse_both) 
```

### Random effects

The random effects of the model give us a ranking of the proportion of models predicting collapse for each assessment unit, regardless of the method used:

```{r}
rr <- ranef(mod_collapse_both)
rr$cond$unit %>% arrange(desc(`(Intercept)`))
```

### EMmeans

```{r}
library(emmeans)
emmeans(
  mod_collapse_both, 
  ~ method | scenario + time, 
  type="response", 
  allow.new.levels=TRUE)
```

