---
title: "Temporal change in Relative Severity"
output: html_document
editor_options: 
  chunk_output_type: console
---


We calculate temporal change in Relative Severity by applying the RS formula to one fixed initial value and a sequence of final values, the final values can represent the state of the indicator in consecutive years (time series), or in different periods (temporal sequence). 

```{r}
#| label: load libraries and functions
#| eval: true
#| echo: false
#| message: false
library(dplyr)
library(ggplot2)
library(units)
library(stringr)
library(ggrepel)
library(purrr)
library(readr)
library(tidyr)
library(forcats)

here::i_am("docs-src/RS/time-series.qmd")
source(here::here("inc","R","RS-functions.R"))
source(here::here("inc","R","RS-shared.R"))
```

```{r}
#| label: read input data 
#| eval: true
#| message: false
target_dir <- "sandbox"
results_file <- here::here(target_dir, "massbalance-model-data-all-groups.rds")
massbalance_results <- readRDS(results_file)

rds_file <- here::here(target_dir,"massbalance-totalmass-all-groups.rds")
totalmass_year_data <- readRDS(rds_file)
results_file <- here::here(target_dir, "relative-severity-degradation-suitability-all-tropical-glaciers.csv")
RS_results <- read_csv(results_file, show_col_types = FALSE) %>%
  mutate(unit_name=str_replace_all(unit," ","-"))
```



## Extent of decline for loss of ice mass

Given the nature of ice mass as a prediction of a physical measure, we need to weight the RS values of individual glacier outlines. 

If we look at the example of the T.G.E. Kilimanjaro, the value of $RS$ for one combination of model and scenario and the year 2040, we have following values of RS:

```{r}
mass_su <- massbalance_results %>%
  filter(
    unit_name %in% c("Kilimanjaro"),
    model_nr %in% c(" 9"),
    scn %in% c("ssp126")
  ) %>%
  mutate(min_mass = mass-mad, max_mass = mass+mad) 

wgs <- mass_su %>%
    filter(year==2000) %>%
    transmute(RGIId,initial_mass=mass) 
 wgs <- wgs   %>%
    mutate(w=initial_mass/sum(wgs$initial_mass))

RSvals <- mass_su %>%
  select(RGIId,year,mass,min_mass,max_mass) %>%
  group_by( RGIId) %>%
  group_modify( ~ RSts(
    .x$year,
    .x$mass,
    vmin = .x$min_mass,
    vmax = .x$max_mass,
    formula = "conditional"
    )
    ) %>%
    left_join(wgs, by = c( "RGIId")) 

d <- RSvals %>%
    filter(year %in% 2040)
d %>% select(RS, initial_mass, w) 
 #%>%
 #   arrange(RS) %>%
 #   ungroup %>% select(w) %>%
 #   map(~cumsum(.x))
```

We can use these values to calculate $ED_{RS≥x}$ as a weighed empirical distribution function:

```{r}
##library(Hmisc)
##ED <- Ecdf(d$RS, what="1-f", weights=d$w, normwt = TRUE, pl=FALSE)
#library(spatstat.geom)
#plot(e <- ewcdf(d$RS, d$w, normalise=TRUE))
 
ED <- ED_w(RS=d$RS,w=d$w)
f <- cED_w(RS=d$RS,w=d$w)
xvals <- seq(0,1,length=20)
ED2 <- tibble(x=xvals,cED=f(xvals)) 

AUC_ED <- integrate(cED_w(RS=d$RS,w=d$w), 0, 1)

ggplot() + 
  #geom_step(data=data.frame(ED),
  #  aes(x,y,col="Hmisc")) +
  geom_step(data=ED, aes(x=x,y=ED)) +
  geom_step(data=ED2, aes(x=x,y=cED),col=2) +
  ylab(expression(ED[RS>=x])) +
  ylim(0,1)+
  xlab("x")
```

We then calculate $ED_{RS≥x}$ for multiple years.

```{r}
d <- RSvals %>%
    filter(year %in% c(2040,2070,2100))

EDs <- d %>%
  group_by(year) %>%
  group_modify( ~ ED_w(
    RS=.x$RS,
    w=.x$w))
ggplot(EDs) + 
  geom_step(aes(x=x, y=ED, group=year,colour=factor(year))) +
  ylab(expression(ED[RS>=x])) +
  ylim(0,1)+
  xlab("x")

```

This figure shows a large increase in values of $RS≥0.5$ between 2040 and 2070 and then an increase of values of $RS≥0.75$ between 2070 and 2100.


 Example from Ecuador:
 We see from the mean RS values that there is an expected increase from 0.487 (VU) to 0.746 (EN).

```{r}
RS_results %>%
  filter(
    threshold=="acc",
    unit %in% "Ecuador",
    modelname == "mri-esm2-0",
         pathway == "ssp126"
         ) %>%
  group_by(timeframe, pathway) %>%
  summarise(
    n=n(),
    mean_RS=mean(RS,na.rm=T),
    mean_RS_cor=mean(RS_cor,na.rm=T)) %>%
  DT::datatable()

```

This arrangement allows to read the cummulative extent by different thresholds of relative severity. For example, there ecosystem extent with $RS ≥ 80%$ increases from less than 2% in the first timeframe to 37.2 and 29.1 %, and the extent with $RS ≥ 30%$ increases from just below 80% to more than 96%. 

We can actually visualise the relationship between this curve and the IUCN category threshold in a convenient way:


```{r}
#| label: fig-ecdf-ecuador
#| fig-cap: Comparison of the modified empirical cumulative distribution function of the $RS_{cor}$ values for the Tropical Glacier Ecosystem of Ecuador with the thresholds for IUCN categories of threat. 
plot_b +
  geom_rect(data=RS_extent_combs, 
            aes(fill = category, 
                ymin = RS_min, ymax = RS_max,
                xmin = extent_min, xmax = extent_max),
            alpha=0.44)

```

We see that the top curve overlaps with the threshold for the category VU, and the two lower curves overlap both VU and EN.



## Cordilleras-Norte-de-Peru

The last plot above allows a more compact comparison of different scenarios and timeframes, for example this plot:

```{r}
#| label: fig-hist-norte-peru
#| fig-cap: Histogram of the $RS_{cor}$ values for the Tropical Glacier Ecosystem of Ecuador using the maximum accuracy threshold.
d <- RS_results %>%
  filter(threshold=="acc", 
         unit=="Cordilleras-Norte-de-Peru", 
         modelname == "mri-esm2-0")
plot_a <- 
  ggplot(d) +
    geom_histogram(aes(x=RS_cor*100,fill=IUCN_cat)) +
    facet_grid(pathway~timeframe) +
    scale_fill_brewer(palette = "Oranges",type="seq", direction=-1) +
#    scale_fill_manual(values = IUCN_cat_colours) +
    theme(legend.position = "none") +
    xlab("Relative severity") +
    ylab("Nr. of cells") 
plot_a 
```
this is probably an error? not including all clusters?

```{r}
tt <- RS_results %>%
  filter(threshold=="acc", 
         unit=="Cordilleras-Norte-de-Peru", 
         modelname == "mri-esm2-0",
         pathway == "ssp126") %>%
  pull(IUCN_cat) %>% table

tt

tt/sum(tt)

# RS_results %>% filter(timeframe=="2041-2070",threshold=="acc") %>% arrange(desc(RS_cor)) %>% select(RS,RS_cor)

```

## Cordilleras-Orientales-de-Peru-y-Bolivia
this is the less threatened, shows a transition from VU for short time frames to EN in longer timeframes, histogram almost bimodal.

```{r}
#| eval: false

RSecdf <- tibble()
for (tfm in c("2011-2040", "2041-2070", "2071-2100")) {
  for (thr in c("acc")) {
    RSvals <- RS_results %>%
      filter(
        unit == "Cordilleras-Orientales-de-Peru-y-Bolivia",
        timeframe==tfm,
        threshold==thr) %>%
      pull(RS_cor)
    RSecdf <- RSecdf %>%
      bind_rows(
        {RSvExt(RSvals) %>% mutate(timeframe = tfm, threshold = thr)}
      )
  }
}




ggplot(RSecdf) +
  geom_rect(data=RS_extent_combs, 
            aes(fill = category, 
                ymin = RS_min, ymax = RS_max,
                xmin = extent_min, xmax = extent_max),
            alpha=1) +
  geom_line(aes(x=Extent,y=RS,lty=timeframe)) +
  scale_fill_manual(values = IUCN_cat_colours) +
  labs(title="Cordilleras orientales de Peru y Bolivia", 
    subtitle="Maximum accuracy threshold") +
    xlab("Cummulative extent") +
    ylab("Relative severity")
```


